Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2017-08-20 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
KM Comments & CAF Figures JPG

A range of methods used to simulate characteristics of fires were applied using data from the source fire experiments as input.  The algebra-based predictions for heat flux to a target under-predicted the measured values.  The gasoline fueled fires exhibited best agreement of 22\% or less for the heat flux predictions.  The polyurethane foam results exhibited the worst agreement with a difference of at least 40\% between the measured and calculated values.  Improved understanding of the radiative fraction from different fuels would improve the capability of the predictions.

@@

This section reviews the fire problem to demonstrate the need and value to society that fire investigation can provide. Following the review, a brief presentation on the analysis  of fire patterns, the use of analytical tools, and the current challenges to the field of fire forensics is given because they are the reason for this study.  At the end of this chapter, an example of a fire incident is presented. In Chapter 9, we will return to the example again.

@@

In the United States, the loss of life due to unwanted fire has been decreasing steadily since the 1970’s.  (This trend does not include loss of life as a result of the terrorist attack on September 11, 2001.)  The estimated number of fire deaths in the U.S. in 1975 was 8,100, or 3.75 deaths per 100,000 people~\cite{America_Burning_Revisited}.  In 2013, there were 3,240 civilian fire fatalities, or 1.02 deaths per 100,000 people~\cite{Karter:2014}.  During this same period, the number of fires in the United States has decreased from 2.5 million fires to 1.24 million fires.  There are many reasons that led to this decrease in loss of life due to fire in the United States.  One example is research and data that enabled the development of product safety standards, improved building codes, improved fire safety systems, and fire prevention programs.  None of this happened without the constant effort of agencies at the federal, state, and local levels working with professional fire organizations in order to deliver the five E’s of fire prevention: engineering, education, enforcement, economic Incentive, and emergency Response~\cite{FEMA:2013}.

@@

The statistics cited above are believed to be the best available in the United States, which maintains perhaps one of the better fire incident databases in the world.  However, an investigative report by Icove and Hargrove indicates arson fires are underreported to the U.S. National Fire Incident Reporting System (NFIRS).  The investigation provided examples from several major cities, such as New York City. In 2011, 11 arson fires in buildings were reported to NFIRS, while the Fire Department of the City of New York's Bureau of Fire Investigation recorded 1,347 arson fires in buildings that year.  Similar discrepancies were found in eight other large U.S. cities.  On average, three-fourths of the arsons uncovered by city fire investigators went unreported to NFIRS~\cite{Icove_2014}.  In other words, the number of arson-related incidents is likely to be under represented in the fire incident data.

@@

\subsection{Fire Investigation Under Review}
The fire investigation community recognized a need to improve the science and practice of fire investigation.  One of the manifestations of the efforts to improve the practice was the development of NFPA~921, {\em Guide for Fire and Explosion Investigations}~\cite{NFPA:921}.  The first edition of NFPA~921 was issued in 1992.  As provided by the document scope, it ``is designed to assist individuals who are charged with the responsibility of investigating and analyzing fire and explosion incidents and rendering opinions as to the origin, cause, responsibility, or prevention of such incidents, and the damage and injuries which arise from such incidents.'' The document’s purpose includes the goal to be ``a model for the advancement and practice of fire and explosion investigation, fire science, and methodology.'' However, NFPA~921 can only incorporate the fire science and research results that have been produced.  Although NFPA~921 is a guide as opposed to a standard, it is considered the standard of care for the fire investigation community in countries that recognize and use NFPA standards and guides.

@@

Another landmark document regarding the state of forensic science in the United States was published in 2009.  The 328 page report, entitled {\em Strengthening Forensic Science in the United States: A Path Forward}, by the U.S.~National Academy of Sciences, reviewed 13 different forensic science disciplines including biological evidence, friction ridge patterns, tool mark and firearms identification, and analysis of explosives evidence and fire debris.  The review of the analysis of explosives evidence and fire debris is covered in three pages of the report.  The summary assessment of analysis of explosives evidence is positive, ``the scientific foundations exist to support the analysis of explosions, because such analysis is based primarily on well-established chemistry''~\cite{Forensic:2009}.  The complete summary assessment for fire debris analysis states: ``By contrast, much more research is needed on the natural variability of burn patterns and damage characteristics and how they are affected by the presence of various accelerants.  Despite the paucity of research, some arson investigators continue to make determinations about whether or not a particular fire was set.  However, according to testimony presented to the committee, many of the rules of thumb that were typically assumed to indicate that an accelerant was used (e.g., alligatoring of wood, specific char patterns) have been shown not to be true.  Experiments should be designed to put arson investigations on a more solid scientific footing.''
This study is in response to the assessment above, specifically regarding the natural variability of fire patterns.

@@

The majority of intentional fires in structures have occurred in residential occupancies, so this study focused on fire patterns on painted gypsum wallboard, which is a common interior finish for residential occupancies.  A common height for a residential occupancies is 2.4~m.  With a wall height of 2.4~m, these experiments used small fires with flame heights of approximately 1.2~m.  The resulting patterns on the gypsum board wall were be representative of a pre-flashover fire pattern.

@@

As you begin to examine the scene, you walk around the exterior and note that on the front of the end unit townhouse, there is fire damage on the exterior above the front door opening that extends across the ceiling of the porch.  There is some thermal damage and smoke deposits on the front siding above the porch.  The outer pane of the living room window is, for the most part, still intact. You notice that three of the four windows on the second floor have been removed.  On side D of the town home, there are no openings and no smoke or fire damage is present.  On the rear or side C of the house, the door and windows are intact but open.  There are no smoke stains and no thermal damage around any of the openings on the rear. No sign of smoke alarms in the home.

You take some notes and photograph the exterior of the structure.  The fire department is repacking their equipment and is preparing to leave when you enter through the front doorway.  The living room has extensive fire damage, as shown in Figure~\ref{EXT_INTLR}.  Looking to your right, two chairs in front of the living room window have some thermal damage.  There is a clear horizontal line of demarcation on the large wood TV stand in the corner.  As your eyes move from the corner back toward the rear of the town home and the kitchen, you notice that the elevation of the line of demarcation is increasing.  This may be an indicator that the origin of the fire is closer to the front of the living room.  Between the two wood cabinets on the D wall of the living room there is a wood framed sofa.  The cushions show thermal damage from pyrolysis.  It is not clear that they were on fire because there is no thermal damage on the wall directly above the top of the sofa cushions that would indicate the presence of a thermal plume.  Still standing in the entry doorway, you turn your attention to the remains of an upholstered sofa and chair.  The wall behind the sofa and chair has extensive thermal damage.  The wall extends from the opening to the dining/kitchen area of the house to the open stairway that goes to the upper level.  You move to the dining area and kitchen area.  There is plenty of smoke damage high in the space, but there are no indications of fire damage due to combustion.  You return to the living room and walk up the stairs.  There is evidence of paint and paper burned off of the gypsum wallboard in the stairway.  The thermal damage does not extend beyond the top of the stairs.  There is smoke damage in the upstairs hallway and in the front bedrooms. The further you move from the stairs, the less thermal damage you find.

You head back down the stairs to further examine the living room, which appears to be the room of fire origin. Your focus on the area with the most damage, the area with the sofa and chair.  You think this may be the area of origin.

As you look toward the B wall of the living room, you note that the right side of the chair is badly burned, and burned from the floor level up.  The same is true from the left side of the sofa.  In fact, the left side of the sofa has more fire damage to its wood frame than the right side.  The corner formed by the wall behind the chair and the wall of the stairway has a V pattern burned into it from the floor and extending up toward the ceiling.  As you look at the debris surrounding the chair, you notice the remains of an electric heater.  The burned wire from the heater extends toward the wall outlet behind the sofa. Figure~\ref{EXT_LRDetails} shows some of these details.

@@

A series of experiments were conducted to characterize a limited set of  residential scale, pre-flashover/early growth stage fires.  A fixed fire source footprint, 30.5~cm square, was chosen for this study, and a range of common fuels that could be found in and around a residence were selected.  The fires were positioned under an oxygen consumption calorimetry hood to determine the heat release rate.  Heat release rate, flame/plume gas temperatures, and total heat flux from the flames were measured.  In addition, flame movement and height were recorded with photographs and videos.  Given the focus on residential fires, small source fires were selected so that the burn pattern would be limited to less than 2.4~m in height.  Replicate experiments were conducted with each of the fuels in order to examine the repeatability of the fires and to quantify the variability in the flames.

@@

The gasoline pan fire used 500~mL of unleaded, 87 octane gasoline.  The pan was constructed from 6~mm thick steel.  The inner dimensions of the pan were 30.5~cm on each side and 1.9~cm high.  The pan was elevated in order to bring the pan lip height up to 7.5~cm above the floor to be the same height as the natural gas burner.  The 500~mL of gasoline had a fuel depth of 5.5~mm and provided a burn time of approximately 300~s.

Flexible polyurethane foam was chosen as the solid fuel. According to the Polyurethane Foam Association, flexible polyurethane foam is the most widely used cushioning material in upholstered furnishings and mattresses. More than 1.2 billion pounds are produced in the U.S. every year~\cite{Foam,Polyurethane_Foam}. In studies by Ahrens and Evarts of the NFPA, upholstered furniture, and mattresses and bedding were the first items ignited in 33\% of the fatal fires in the U.S. based on averaged data from 2005 through 2009~\cite{Ahrens:2011,Evarts:2011}.

@@

The source fire characterization experiments were conducted in the NIST Large Fire Research Laboratory's 3~m x 3~m hood oxygen consumption calorimeter. This apparatus follows the methodology of ASTM~E 2067-08, {\em Standard Practice for Full-Scale Oxygen Consumption Calorimetry Fire Tests} ~\cite{ASTM_E2067}. The fires were centered under the hood of the calorimeter and positioned in an open steel frame, which provided support for the thermocouples and heat flux sensors. A load cell was positioned on the floor as a support for the fuels.  All of the data was recorded at intervals of 1~s on a computer-based data acquisition system.

@@

The NIST Large Fire Research Laboratory's 3~m x 3~m oxygen depletion calorimeter had an estimated peak heat release rate capacity of approximately 700~kW.  Details on the operation and uncertainty in measurements associated with NIST's 6~m x 6~m oxygen consumption calorimeter were documented by Bryant et al.~\cite{Bryant:2004}. Many of the same instruments were shared between the two calorimeters and calibration burns were conducted during each experimental series.  As a result, the expanded uncertainty of this device has been estimated as 11\% on the measured heat release rate.

@@

Gas temperatures were measured with exposed-bead, Chromel-Alumel (type K) thermocouples, with a 1.0~mm nominal diameter.  Starting from the exposed bead, the thermocouple wire was sheathed in a 3.2~mm diameter inconel shield, 0.76~m in length. The thermocouple was ungrounded. The standard uncertainty in temperature of the thermocouple wire itself is 2.2$^\circ$C at 277$^\circ$C and increases to 9.5$^\circ$C at 871$^\circ$C as determined by the wire manufacturer.  The response time constant is approximately 0.2~seconds~\cite{Omega}.  The variation of the temperature in the environment surrounding the thermocouple was known to be much greater than that of the wire uncertainty~\cite{Blevins:1999,Pitts:2001}. Small diameter thermocouples were used to limit the impact of radiative heating and cooling.  The estimated combined expanded uncertainty for temperature in these experiments was 15\%.

@@

The fires were centered under the hood of the calorimeter on a non-combustible platform measuring 0.91~m x 1.2~m x 12~mm thick supported by the load cell.  The load cell had a resolution of 1.0~g~\cite{Mettler}. The expanded uncertainty was estimated to be 5\%.  The load cell was checked between experiments with calibrated weights during the zero and span checks of the calorimeter.

@@

The natural gas flow to the burner was measured in terms of volumetric flow with a dry test meter that was corrected for temperature and pressure. The flow rate was determined to be 2.1~liters per second.  This yielded a mass flow of natural gas of approximately 2~g/s.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{../Figures/Schematic_Drawing}\\
  \caption[Diagram of the source fire flame and plume instrumentation apparatus]{Diagram of the source fire flame and plume instrumentation apparatus, which was centered under the 3~m x 3~m oxygen consumption calorimeter.  The heat flux gauges were facing toward the center of the burner/fuel area.}
  \label{Drawing_Fire_Flamer}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{../Figures/Source_Fire_Flamer}\\
  \caption[A photograph of the source fire flame and plume instrumentation apparatus]{A photograph of the source fire flamer and plume instrumentation apparatus, which was centered under the 3~m x 3~m hood of the oxygen consumption calorimeter.  In the foreground from left to right are the video, 35~mm SLR, and thermal imaging cameras.}
  \label{Fire_Flamer}
\end{figure}

@@

The block of polyurethane foam demonstrated the largest variability of the three fuels in terms of peak heat release rate and time to peak heat release. Similar to the gasoline that began with a fixed quantity of fuel which was ignited and burned to completion, the polyurethane foam also generated bell shaped heat release rate curve.  Because the duration of the heat release rate peaks were less than those of the gasoline, the peak heat release rate averaging was performed over a 15~second interval bounding the peak heat release rate from each of the 10 experiments.  This process yielded an average peak heat release rate of 45~kW~$\pm$~21~kW, or 47\%.  The total heat release over the 300~second period after ignition was 4.1~MJ~$\pm$~0.2~MJ.

@@

As noted by Beyler in his review of fire plume correlations, video tape analysis seems to provide the best estimates of flame height~\cite{Beyler:1986}. Each experiment was recorded with a digital video camera at a refresh rate frequency of 29.97~fps, which is the National Television Standards Committee (NTSC) standard. The resulting image in each frame was 720 pixels wide x 480 pixels high. Segments of the video which corresponded to the peak heat release of each experiment were captured with a video editing program.

@@

A free standing steel frame wall with a 12.7~mm thick, 2.44~m high x 1.22~m wide calcium silicate board face was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux. Marinite Type I is the commercial name of the non-combustible board used in these experiments.  It is normally used as a fire-resistant thermal insulation that also has structural strength. 90\% of the board is composed of calcium silicate and calcium meta-silicate.  The remaining 10\% of the board is composed of a combination of organic fibers, fiber glass filament and crystalline silica (quartz)~\cite{BNZ:MSDS}.  Material properties that impact heat transfer were selected from the manufacturer's literature and are presented In Table~\ref{tab:Marinite_Thermal_Properties}~\cite{Marinite:1997}.

@@

Two vertical arrays of water cooled, Schmidt-Boelter total heat flux gauges, as previously described in the instrumentation section above, were installed at 20 cm intervals from 0.2~m to 1.2~m above the burner.  One vertical array was centered on the burner and the other vertical array was centered on the right edge of the burner. The arrangement is shown in Fig.~\ref{Instrumented_NC_Wall}. Bare-bead, Chromel-Alumel (type K) thermocouples with a 0.5~mm nominal diameter were installed next to the center of each heat flux sensor and continued, at 20~cm intervals, up to a height of 2.2~m above the surface of the burner. Photographs of the front and rear of the instrumented wall are shown in Figures~\ref{Instrumented_Wall_Front_photo} and \ref{Instrumented_Wall_Rear_photo}. The thermocouples were installed in small holes from the back of the wall.  The bead of the thermocouple was flush with the surface on the calcium silicate board and was in contact with the board, as shown in Figure~\ref{Instrumented_Wall_Close_up_TC_HF_photo}.  For the center array, a second thermocouple was installed on the backside of the calcium silicate board, approximately 10~mm below the thermocouple positioned on the front surface of the wall. The photograph in Figure~\ref{Instrumented_Wall_Detail_Rear_TC_photo} shows a rear thermocouple with the bead in contact with the rear surface of the wall and the relative position to the thermocouple on the front surface of the wall. In a manner similar to the source fire characterization experiments, the instrumented wall experiments were conducted under the NIST Large Fire Research Laboratory's 3~m x 3~m oxygen depletion calorimeter.

@@

Figures~\ref{TC_Surf_Cent_TWNG_comp} and \ref{TC_Surf_Edge_TWNG_comp} show consistent trends of the thermocouple temperatures in the vertical arrays on the wall increasing as the burner was moved toward the wall.  Figure~\ref{TC_Back_Center_TWNG_comp} shows that the peak temperatures aligned with the vertical center of the burner on the backside of the gypsum wallboard peaked at approximately 100$^\circ$C.

@@

Temperatures were also measured on the backside of the wall along the vertical centerline of the burners, see Figure~\ref{NCTW_Back_Cent_comp}. The temperature difference between the peak average temperatures on the front (fire side) of the wall at 0.2~m above the burner versus the corresponding temperatures on the backside of the wall show a difference of at least 400$^\circ$C.  It is consistent with the temperature data shown in Figures~\ref{NCTW_Surf_Center_comp} and \ref{NCTW_Surf_Edge_comp} that as the burners were moved away from the wall, the temperature differences decreased.

@@

The gypsum wallboard used in these experiments is referred to as ``regular'' as opposed to Type C or Type X gypsum wallboard, which are used in fire-rated assemblies. The gypsum wallboard used in these experiments was purchased from the same manufacturer in an effort to have the most similar panels for use between the experiments. Each 1.22~m x 2.44~m x 12.7~mm thick panel weighed approximately 30.5~kg.  The thermal properties of the gypsum wallboard panels are given in the Table~\ref{tab:Gypsum wallboard_Thermal_Properties}.  The first two rows of data were measured from samples of the gypsum wallboard used in these experiments.  The bottom four rows of data are from the Gypsum Association~\cite{Gross:1985}.

>>>>

A range of methods used to simulate characteristics of fires were applied using data from the source fire experiments as input.  The empirical-based predictions for heat flux to a target under-predicted the measured values.  The gasoline fueled fires exhibited best agreement of 22\% or less for the heat flux predictions.  The polyurethane foam results exhibited the worst agreement with a difference of at least 40\% between the measured and calculated values.  Improved understanding of the radiative fraction from different fuels would improve the capability of the predictions.

@@

This section reviews the fire problem to demonstrate the need and value to society that fire investigation can provide. Following the review, a brief presentation on the analysis  of fire patterns, the use of analytical tools, and the current challenges to the field of fire forensics is given because they are the reason for this study.  At the end of this chapter, an example of a fire incident is presented. In Chapter 9, the example fire incident will be revisited.

@@

In the United States, the loss of life due to unwanted fire has been decreasing steadily since the 1970’s.  (This trend does not include loss of life as a result of the terrorist attack on September 11, 2001.)  The estimated number of fire deaths in the U.S. in 1975 was 8,100, or 3.75 deaths per 100,000 people~\cite{America_Burning_Revisited}.  In 2013, there were 3,240 civilian fire fatalities, or 1.02 deaths per 100,000 people~\cite{Karter:2014}.  During this same period, the number of fires in the United States has decreased from 2.5 million fires to 1.2 million fires.  There are many reasons that led to this decrease in loss of life due to fire in the United States.  One example is research and data that enabled the development of product safety standards, improved building codes, improved fire safety systems, and fire prevention programs.  None of this happened without the constant effort of agencies at the federal, state, and local levels working with professional fire organizations in order to deliver the five E’s of fire prevention: engineering, education, enforcement, economic incentive, and emergency response~\cite{FEMA:2013}.

@@

The statistics cited above are believed to be the best available in the United States, which maintains perhaps one of the better fire incident databases in the world.  However, an investigative report by Icove and Hargrove indicates arson fires are underreported to the U.S. National Fire Incident Reporting System (NFIRS)~\cite{Icove_2014}.  The investigation provided examples from several major cities, such as New York City. In 2011, 11 arson fires in buildings were reported to NFIRS, while the Fire Department of the City of New York's Bureau of Fire Investigation recorded 1,347 arson fires in buildings that year.  Similar discrepancies were found in eight other large U.S. cities.  On average, three-fourths of the arsons uncovered by city fire investigators went unreported to NFIRS~\cite{Icove_2014}.  In other words, the number of arson-related incidents is likely to be under represented in the fire incident data.

@@

\subsection{Fire Investigation Under Review}
The fire investigation community recognized a need to improve the science and practice of fire investigation.  One of the manifestations of the efforts to improve the practice was the development of NFPA~921, {\em Guide for Fire and Explosion Investigations}~\cite{NFPA:921}.  The first edition of NFPA~921 was issued in 1992.  As provided by the document scope, it ``is designed to assist individuals who are charged with the responsibility of investigating and analyzing fire and explosion incidents and rendering opinions as to the origin, cause, responsibility, or prevention of such incidents, and the damage and injuries which arise from such incidents.'' The document’s purpose includes the goal to be ``a model for the advancement and practice of fire and explosion investigation, fire science, and methodology.'' However, NFPA~921 can only incorporate the fire science and research results that have been produced.  Although NFPA~921 is a guide as opposed to a standard, it is considered the \textit{standard of care} for the fire investigation community in countries that recognize and use NFPA standards and guides.

@@

Another landmark document regarding the state of forensic science in the United States was published in 2009.  The 328 page report, entitled {\em Strengthening Forensic Science in the United States: A Path Forward}, by the U.S.~National Academy of Sciences, reviewed 13 different forensic science disciplines including biological evidence, friction ridge patterns, tool mark and firearms identification, and analysis of explosives evidence and fire debris.  The review of the analysis of explosives evidence and fire debris is covered in three pages of the report.  The summary assessment of analysis of explosives evidence is positive, ``the scientific foundations exist to support the analysis of explosions, because such analysis is based primarily on well-established chemistry''~\cite{Forensic:2009}.  The complete summary assessment for fire debris analysis states:

\begin{quotation}
By contrast, much more research is needed on the natural variability of burn patterns and damage characteristics and how they are affected by the presence of various accelerants.  Despite the paucity of research, some arson investigators continue to make determinations about whether or not a particular fire was set.  However, according to testimony presented to the committee, many of the rules of thumb that were typically assumed to indicate that an accelerant was used (e.g., alligatoring of wood, specific char patterns) have been shown not to be true.  Experiments should be designed to put arson investigations on a more solid scientific footing.
\end{quotation}

This study is in response to the assessment above, specifically regarding the natural variability of fire patterns.

@@

The majority of intentional fires in structures in the United States have occurred in residential occupancies, so this study focused on fire patterns on painted gypsum wallboard, which is a common interior finish for residential occupancies.  A common height for a residential occupancies is 2.4~m.  With a wall height of 2.4~m, these experiments used small fires with flame heights of approximately 1.2~m.  The resulting patterns on the gypsum board wall were be representative of a pre-flashover fire pattern.

@@

As you begin to examine the scene, you walk around the exterior and note that on the front of the end unit townhouse, there is fire damage on the exterior above the front door opening that extends across the ceiling of the porch.  There is some thermal damage and smoke deposits on the front siding above the porch.  The outer pane of the living room window is, for the most part, still intact. You notice that three of the four windows on the second floor have been removed.  On the right side of the town home, there are no openings and no smoke or fire damage is present.  On the rear of the house, the door and windows are intact but open.  There are no smoke stains and no thermal damage around any of the openings on the rear.

You take some notes and photograph the exterior of the structure.  The fire department is repacking their equipment and is preparing to leave when you enter through the front doorway.  The living room has extensive fire damage, as shown in Figure~\ref{EXT_INTLR}.  Looking to your right, two chairs in front of the living room window have some thermal damage.  There is a clear horizontal line of demarcation on the large wood TV stand in the corner.  As your eyes move from the corner back toward the rear of the town home and the kitchen, you notice that the elevation of the line of demarcation is increasing.  This may be an indicator that the origin of the fire is closer to the front of the living room.  Between the two wood cabinets on the right wall of the living room there is a wood framed sofa or futon.  The cushions show thermal damage from pyrolysis.  It is not clear that they were actually aflame because there is no thermal damage on the wall directly above the top of the sofa cushions that would indicate the presence of a thermal plume.  Still standing in the entry doorway, you look almost straight ahead to the remains of an upholstered sofa and chair.  The wall behind the sofa and chair has extensive thermal damage.  The wall extends from the opening to the dining/kitchen area of the house to the open stairway that goes to the upper level.  You move to the dining area and kitchen area.  There is plenty of smoke damage high in the space, but there are no indications of fire damage due to combustion.  You return to the living room and walk up the stairs.  There is evidence of paint and paper burned off of the gypsum wallboard in the stairway.  The thermal damage does not extend beyond the top of the stairs.  There is smoke damage in the upstairs hallway and in the front bedrooms. The further you move from the stairs, the less thermal damage you find. There is no sign of smoke alarms in the home.

You head back down the stairs to further examine the living room, which based on the fire damage appears to be the room of fire origin. You focus on the area with the most damage, the area with the sofa and chair.  You think this may be the area of origin.

As you look again at the chair between the stair and the sofa, you note that the right side of the chair is badly burned, and burned from the floor level up.  The same is true for the left side of the sofa.  In fact, the left side of the sofa has more fire damage to its wood frame than the right side.  The corner formed by the wall behind the chair and the wall of the stairway has a V pattern burned into it from the floor and extending up toward the ceiling.  As you look at the debris surrounding the chair, you notice the remains of an electric heater.  The burned wire from the heater extends toward the wall outlet behind the sofa. Figure~\ref{EXT_LRDetails} shows some of these details.

@@

A series of experiments was conducted to characterize a limited set of  residential scale, pre-flashover/early growth stage fires.  A fixed fire source footprint, 30.5~cm square, was chosen for this study, and a range of common fuels that could be found in and around a residence were selected.  The fires were positioned under an oxygen consumption calorimetry hood to determine the heat release rate.  Heat release rate, flame/plume gas temperatures, and total heat flux from the flames were measured.  In addition, flame movement and height were recorded with photographs and videos.  Given the focus on residential fires, small source fires were selected so that the burn pattern would be limited to less than 2.4~m in height.  Replicate experiments were conducted with each of the fuels in order to examine the repeatability of the fires and to quantify the variability in the flames.

@@

The gasoline pan fire used 500~mL of unleaded gasoline with an octane rating of 87, which is the average of the research octane and motor octane numbers.  The pan was constructed from 6~mm thick steel.  The inner dimensions of the pan were 30.5~cm on each side and 1.9~cm high.  The pan was elevated in order to bring the pan lip height up to 7.5~cm above the floor to be the same height as the natural gas burner.  The 500~mL of gasoline had a fuel depth of 5.5~mm and provided a burn time of approximately 300~s.

Flexible polyurethane foam was chosen as the solid fuel. According to the Polyurethane Foam Association, flexible polyurethane foam is the most widely used cushioning material in upholstered furnishings and mattresses. More than 544 million kg are produced in the U.S. every year~\cite{Foam,Polyurethane_Foam}. In studies by Ahrens and Evarts of the NFPA, upholstered furniture, and mattresses and bedding were the first items ignited in 33\% of the fatal fires in the U.S. based on averaged data from 2005 through 2009~\cite{Ahrens:2011,Evarts:2011}.

@@

The source fire characterization experiments were conducted in the NIST Large Fire Research Laboratory's 3~m by 3~m hood oxygen consumption calorimeter. This apparatus follows the methodology of ASTM~E 2067-08, {\em Standard Practice for Full-Scale Oxygen Consumption Calorimetry Fire Tests} ~\cite{ASTM_E2067}. The fires were centered under the hood of the calorimeter and positioned in an open steel frame, which provided support for the thermocouples and heat flux sensors. A load cell was positioned on the floor as a support for the fuels.  All of the data was recorded at intervals of 1~s on a computer-based data acquisition system.

@@

The NIST Large Fire Research Laboratory's 3~m by 3~m oxygen depletion calorimeter had an estimated peak heat release rate capacity of approximately 700~kW.  Details on the operation and uncertainty in measurements associated with NIST's 6~m by 6~m oxygen consumption calorimeter were documented by Bryant et al.~\cite{Bryant:2004}. Many of the same instruments were shared between the two calorimeters and calibration burns were conducted during each experimental series.  As a result, the expanded uncertainty of this device has been estimated as 11\% on the measured heat release rate.

@@

Gas temperatures were measured with exposed-bead, Chromel-Alumel (type K) thermocouples, with a 1.0~mm nominal diameter.  Starting from the exposed bead, the thermocouple wire was sheathed in a 3.2~mm diameter inconel shield, 0.76~m in length. The thermocouple was ungrounded. The standard uncertainty in temperature of the thermocouple wire itself is 2.2~$^\circ$C at 277~$^\circ$C and increases to 9.5~$^\circ$C at 871~$^\circ$C as determined by the wire manufacturer.  The response time constant is approximately 0.2~seconds~\cite{Omega}.  The variation of the temperature in the environment surrounding the thermocouple was known to be much greater than that of the wire uncertainty~\cite{Blevins:1999,Pitts:2001}. Small diameter thermocouples were used to limit the impact of radiative heating and cooling.  The estimated combined expanded uncertainty for temperature in these experiments was 15\%.

@@

The fires were centered under the hood of the calorimeter on a non-combustible platform measuring 0.91~m by 1.2~m by 12~mm thick supported by the load cell.  The load cell had a resolution of 1.0~g~\cite{Mettler}. The expanded uncertainty was estimated to be 5\%.  The load cell was checked between experiments with calibrated weights during the zero and span checks of the calorimeter.

@@

The natural gas flow to the burner was measured in terms of volumetric flow with a dry test meter that was corrected for temperature and pressure. The flow rate was determined to be 2.1~L/s.  This yielded a mass flow of natural gas of approximately 2~g/s.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{../Figures/Schematic_Drawing}\\
  \caption[Diagram of the source fire flame and plume instrumentation apparatus]{Diagram of the source fire flame and plume instrumentation apparatus, which was centered under the 3~m by 3~m oxygen consumption calorimeter.  The heat flux gauges were facing toward the center of the burner/fuel area.}
  \label{Drawing_Fire_Flamer}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{../Figures/Source_Fire_Flamer}\\
  \caption[A photograph of the source fire flame and plume instrumentation apparatus]{A photograph of the source fire flame and plume instrumentation apparatus, which was centered under the 3~m by 3~m hood of the oxygen consumption calorimeter.  In the foreground from left to right are the video, 35~mm SLR, and thermal imaging cameras.}
  \label{Fire_Flamer}
\end{figure}

@@

The block of polyurethane foam demonstrated the largest variability of the three fuels in terms of peak heat release rate and time to peak heat release. Similar to the gasoline that began with a fixed quantity of fuel which was ignited and burned to completion, the polyurethane foam also generated a bell shaped heat release rate curve.  Because the duration of the heat release rate peaks were less than those of the gasoline, the peak heat release rate averaging was performed over a 15~second interval bounding the peak heat release rate from each of the 10 experiments.  This process yielded an average peak heat release rate of 45~kW~$\pm$~21~kW, or 47\%.  The total heat release over the 300~second period after ignition was 4.1~MJ~$\pm$~0.2~MJ.

@@

As noted by Beyler in his review of fire plume correlations, video tape analysis seems to provide the best estimates of flame height~\cite{Beyler:1986}. Each experiment was recorded with a digital video camera at a refresh rate frequency of 29.97~fps, which is the National Television Standards Committee (NTSC) standard. The resulting image in each frame was 720 pixels wide by 480 pixels high. Segments of the video which corresponded to the peak heat release of each experiment were captured with a video editing program.

@@

A free standing steel frame wall with a 12.7~mm thick, 2.44~m high by 1.22~m wide calcium silicate board face was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux. Marinite Type I is the commercial name of the non-combustible board used in these experiments.  It is normally used as a fire-resistant thermal insulation that also has structural strength. 90\% of the board is composed of calcium silicate and calcium meta-silicate.  The remaining 10\% of the board is composed of a combination of organic fibers, fiber glass filament and crystalline silica (quartz)~\cite{BNZ:MSDS}.  Material properties that impact heat transfer were selected from the manufacturer's literature and are presented In Table~\ref{tab:Marinite_Thermal_Properties}~\cite{Marinite:1997}.

@@

Two vertical arrays of water cooled, Schmidt-Boelter total heat flux gauges, as previously described in the instrumentation section above, were installed at 20 cm intervals from 0.2~m to 1.2~m above the burner.  One vertical array was centered on the burner and the other vertical array was centered on the right edge of the burner. The arrangement is shown in Fig.~\ref{Instrumented_NC_Wall}. Bare-bead, Chromel-Alumel (type K) thermocouples with a 0.5~mm nominal diameter were installed next to the center of each heat flux sensor and continued, at 20~cm intervals, up to a height of 2.2~m above the surface of the burner. Photographs of the front and rear of the instrumented wall are shown in Figures~\ref{Instrumented_Wall_Front_photo} and \ref{Instrumented_Wall_Rear_photo}. The thermocouples were installed in small holes from the back of the wall.  The bead of the thermocouple was flush with the surface on the calcium silicate board and was in contact with the board, as shown in Figure~\ref{Instrumented_Wall_Close_up_TC_HF_photo}.  For the center array, a second thermocouple was installed on the backside of the calcium silicate board, approximately 10~mm below the thermocouple positioned on the front surface of the wall. The photograph in Figure~\ref{Instrumented_Wall_Detail_Rear_TC_photo} shows a rear thermocouple with the bead in contact with the rear surface of the wall and the relative position to the thermocouple on the front surface of the wall. In a manner similar to the source fire characterization experiments, the instrumented wall experiments were conducted under the NIST Large Fire Research Laboratory's 3~m by 3~m oxygen depletion calorimeter.

@@

Figures~\ref{TC_Surf_Cent_TWNG_comp} and \ref{TC_Surf_Edge_TWNG_comp} show consistent trends of the thermocouple temperatures in the vertical arrays on the wall increasing as the burner was moved toward the wall.  Figure~\ref{TC_Back_Center_TWNG_comp} shows that the peak temperatures aligned with the vertical center of the burner on the backside of the gypsum wallboard peaked at approximately 100~$^\circ$C.

@@

Temperatures were also measured on the backside of the wall along the vertical centerline of the burners, see Figure~\ref{NCTW_Back_Cent_comp}. The temperature difference between the peak average temperatures on the front (fire side) of the wall at 0.2~m above the burner versus the corresponding temperatures on the backside of the wall show a difference of at least 400~$^\circ$C.  It is consistent with the temperature data shown in Figures~\ref{NCTW_Surf_Center_comp} and \ref{NCTW_Surf_Edge_comp} that as the burners were moved away from the wall, the temperature differences decreased.

@@

The gypsum wallboard used in these experiments is referred to as ``regular'' as opposed to Type C or Type X gypsum wallboard, which are used in fire-rated assemblies. The gypsum wallboard used in these experiments was purchased from the same manufacturer in an effort to have the most similar panels for use between the experiments. Each 1.22~m by 2.44~m by 12.7~mm thick panel weighed approximately 30.5~kg.  The thermal properties of the gypsum wallboard panels are given in the Table~\ref{tab:Gypsum wallboard_Thermal_Properties}.  The first two rows of data were measured from samples of the gypsum wallboard used in these experiments.  The bottom four rows of data are from the Gypsum Association~\cite{Gross:1985}.

A number of studies have examined gypsum wallboard under elevated temperature conditions in terms of either its thermal properties or its physical change as a result of exposure to heat.  Schroeder conducted a study on the post-fire analysis of construction materials, including gypsum.  As gypsum is heated to temperatures above 80~$^\circ$C it begins to change phase and degrade.  Continued heating causes some of the chemically bound water to disassociate, and the gypsum changes to a hemihydrate.  This transformation occurs in the range of 98~$^\circ$C to 130~$^\circ$C. With additional heating, the residue of the gypsum continues to transform and soften, turning into anhydrite, a process commonly known as calcination.  Schroeder exposed samples of gypsum to a range of heat fluxes and temperatures to determine the time for 80~$^\circ$C (dihydrate phase), 200~$^\circ$C (hemihydrate phase), and 500~$^\circ$C (anhydrous phase) isotherms to penetrate to different depths in the gypsum.  The crystal phase and the depth of penetration of the phase change was determined with the use of X-ray diffraction~\cite{Schroeder:1999}.  NFPA 921 also discusses calcination and how an investigator can use the visual appearance or probing of the soft dehydrated gypsum to map fire patterns.
